---
title: "Meta-analysis METAL"
author: "daniel_linares"
output: html_document
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

shell_call <- function(command, ...) {
  result <- system(command, intern = TRUE, ...)
  cat(paste0(result, collapse = "\n"))
}
```


```{r}
install.packages("R.utils")
library(data.table)
library(tidyverse)
```


## Install METAL

```{r cars}
shell_call("cd")
shell_call("curl -O https://csg.sph.umich.edu/abecasis/Metal/download/Linux-metal.tar.gz")
shell_call("tar -xvzf Linux-metal.tar.gz")
```

```{r}
shell_call("sudo mv generic-metal/metal /usr/local/bin/.")
```

```{r}
shell_call("curl -O https://de.cyverse.org/dl/d/9DEF6A65-F0A7-4C6E-8633-957C118FD1B3/meta_GWAS.tar.gz")
```

```{r}
shell_call("tar -xvzf meta_GWAS.tar.gz")
```

```{r}
shell_call("metal")
```

# Meta-analysis of Congenital malformations of Great Arteries

## Importing GWAS Datasets

### Import Congenital malformations of Great Arteries from FinnGen

```{r pressure, echo=FALSE}
shell_call("curl https://storage.googleapis.com/finngen-public-data-r12/summary_stats/release/finngen_R12_Q17_CONGEN_MALFO_GREAT_ARTERIES.gz -o finngen_R12_Q17_CONGEN_MALFO_GREAT_ARTERIES.gz")

```

```{r}
chd_finngen <-
  fread("/content/finngen_R12_Q17_CONGEN_MALFO_GREAT_ARTERIES.gz")
```

```{r}
chd_finngen[, N := 448451 + 687]
```

```{r}
chd_finngen <- chd_finngen %>%
  select(SNP = rsids,
         CHR = '#chrom',
         A1 = alt, A2 = ref, N, B = beta, SE = "sebeta", P = pval)
```

```{r}
#chd_finngen <- chd_finngen[snp_adhd, on = "SNP", nomatch = 0]
```

```{r}
fwrite(chd_finngen,
  file =
    "chd_summaries.FinnGen.txt",
  sep = "\t"
)
```

```{r}
remove(chd_finngen)
```

### Import Congenital malformations of Great Arteries from UK BB

```{r}
shell_call("curl https://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST90474001-GCST90475000/GCST90474263/harmonised/GCST90474263.h.tsv.gz -o GCST90474260.h.tsv.gz")
```

The file is to large for Golab, the data manipulation was done in a different way XD

```{r}
system("zcat /content/GCST90474260.h.tsv.gz | awk -F'\t' '{print $1,$3,$4,$5,$6,$8,$9,$12}' OFS='\t' | gzip > reduced_file_ukbb.tsv.gz")
```

```{r}
system("gunzip -c /content/reduced_file_ukbb.tsv.gz > reduced_file_ukbb.tsv")
```

Here I import the ADHD GWAS from PGC to select the necesary SNPs, reducing the UKBB to only relevant SNPs

```{r}
shell_call('wget --user-agent="Mozilla/5.0 (X11; Linux x86_64) AppleWebKit/537.36 Chrome/120.0" -O adhdPGC.zip "https://figshare.com/ndownloader/articles/22564390/versions/1"')
```

```{r}
shell_call("unzip /content/adhdPGC.zip")
```

```{r}
snp_adhd <- fread("/content/ADHD2022_iPSYCH_deCODE_PGC.meta.gz", select = c("SNP"))
```

```{r}
fwrite(data.table(SNP = snp_adhd$SNP), "snps_to_keep.txt", col.names = FALSE)
```

```{r}
# Filtering
system("awk 'NR==1{print; next} FNR==NR{a[$1]; next} $7 in a' snps_to_keep.txt /content/reduced_file_ukbb.tsv > filtered_file_ukbb.tsv")
```

```{r}
chd_ukbb <- fread("filtered_file_ukbb.tsv")
```

```{r}
colnames(chd_ukbb) <- c("chromosome", "effect_allele",	"other_allele",	"beta",	"standard_error",	"p_value",	"rsid",	"n")
```

```{r}
fwrite(chd_ukbb,
  file =
    "chd_summaries.UKBB.txt",
  sep = "\t"
)
```

```{r}
remove(chd_ukbb)
```

## Run METAL Great Arteries

```{r}

system("apt-get update -qq && apt-get install -y build-essential cmake git wget -qq")


if (!dir.exists("METAL")) {
  system("git clone https://github.com/statgen/METAL.git METAL")
}

system("mkdir -p METAL/build && cd METAL/build && cmake .. >/dev/null && make -j2 >/dev/null", wait = TRUE)


possible_paths <- c("METAL/build/metal", "METAL/metal", "METAL/build/metal/metal")
metal_path <- Sys.which("metal")  # if in PATH
if (metal_path == "") {
  metal_path <- possible_paths[file.exists(possible_paths)][1]
}
if (is.na(metal_path) || metal_path == "") stop("METAL binary not found. Check build logs in METAL/build.")

cat("METAL binary found at:", metal_path, "\n")
```


```{r}
metal_script <- c(
  "# METAL meta-analysis commands",
  "SCHEME STDERR",

  "# READ in FinnGen data",
  "MARKER SNP",
  "ALLELE A1 A2",
  "EFFECT B",
  "STDERR SE",
  "PVALUE P",
  "WEIGHT N",
  "PROCESS /content/chd_summaries.FinnGen.txt",

  "# READ in UK Biobank data",
  "MARKER rsid",
  "ALLELE effect_allele other_allele",
  "EFFECT beta",
  "STDERR standard_error",
  "PVALUE p_value",
  "WEIGHT n",
  "PROCESS /content/chd_summaries.UKBB.txt",

  "ANALYZE"
)
writeLines(metal_script, con = "metal_script.txt")
```

```{r}

log_out <- "metal_run.log"
log_err <- "metal_run.err"
cmd <- sprintf("%s metal_script.txt > %s 2> %s", shQuote(metal_path), log_out, log_err)
system(cmd, wait = TRUE)


cat("=== METAL STDOUT (tail) ===\n")
system(sprintf("tail -n 50 %s || true", log_out))
cat("\n=== METAL STDERR (tail) ===\n")
system(sprintf("tail -n 50 %s || true", log_err))


list.files(pattern = "META|meta_results|METAANALYSIS", recursive = FALSE)
```

```{r}
shell_call("mv METAANALYSIS1.TBL METAANALYSIS1_Great.TBL")
```

# Meta-analysis of Septal Defects

## Importing GWAS Datasets

### Import Congenital malformations of Great Arteries from FinnGen

```{r pressure, echo=FALSE}
shell_call("curl https://storage.googleapis.com/finngen-public-data-r12/summary_stats/release/finngen_R12_Q17_SEPTA_DEFEC.gz -o finngen_R12_Q17_SEPTA_DEFEC.gz")

```

```{r}
chd_finngen <-
  fread("/content/finngen_R12_Q17_SEPTA_DEFEC.gz")
```

```{r}
chd_finngen[, N := 448451 + 687]
```

```{r}
chd_finngen <- chd_finngen %>%
  select(SNP = rsids,
         CHR = '#chrom',
         A1 = alt, A2 = ref, N, B = beta, SE = "sebeta", P = pval)
```

```{r}
#chd_finngen <- chd_finngen[snp_adhd, on = "SNP", nomatch = 0]
```

```{r}
fwrite(chd_finngen,
  file =
    "chd_summaries.FinnGen.txt",
  sep = "\t"
)
```

```{r}
remove(chd_finngen)
```

### Import Congenital malformations of Great Arteries from UK BB

```{r}
shell_call("curl https://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST90474001-GCST90475000/GCST90474260/harmonised/GCST90474260.h.tsv.gz -o GCST90474260.h.tsv.gz")
```

The file is to large for Golab, the data manipulation was done in a different way XD

```{r}
system("zcat /content/GCST90474260.h.tsv.gz | awk -F'\t' '{print $1,$3,$4,$5,$6,$8,$9,$12}' OFS='\t' | gzip > reduced_file_ukbb.tsv.gz")
```

```{r}
system("gunzip -c /content/reduced_file_ukbb.tsv.gz > reduced_file_ukbb.tsv")
```

Here I import the ADHD GWAS from PGC to select the necesary SNPs, reducing the UKBB to only relevant SNPs

```{r}
shell_call('wget --user-agent="Mozilla/5.0 (X11; Linux x86_64) AppleWebKit/537.36 Chrome/120.0" -O adhdPGC.zip "https://figshare.com/ndownloader/articles/22564390/versions/1"')
```

```{r}
shell_call("unzip /content/adhdPGC.zip")
```

```{r}
snp_adhd <- fread("/content/ADHD2022_iPSYCH_deCODE_PGC.meta.gz", select = c("SNP"))
```

```{r}
fwrite(data.table(SNP = snp_adhd$SNP), "snps_to_keep.txt", col.names = FALSE)
```

```{r}
# Filtering
system("awk 'NR==1{print; next} FNR==NR{a[$1]; next} $7 in a' snps_to_keep.txt /content/reduced_file_ukbb.tsv > filtered_file_ukbb.tsv")
```

```{r}
chd_ukbb <- fread("filtered_file_ukbb.tsv")
```

```{r}
colnames(chd_ukbb) <- c("chromosome", "effect_allele",	"other_allele",	"beta",	"standard_error",	"p_value",	"rsid",	"n")
```

```{r}
fwrite(chd_ukbb,
  file =
    "chd_summaries.UKBB.txt",
  sep = "\t"
)
```

```{r}
remove(chd_ukbb)
```

## Run METAL Great Arteries

```{r}

system("apt-get update -qq && apt-get install -y build-essential cmake git wget -qq")


if (!dir.exists("METAL")) {
  system("git clone https://github.com/statgen/METAL.git METAL")
}

system("mkdir -p METAL/build && cd METAL/build && cmake .. >/dev/null && make -j2 >/dev/null", wait = TRUE)


possible_paths <- c("METAL/build/metal", "METAL/metal", "METAL/build/metal/metal")
metal_path <- Sys.which("metal")  # if in PATH
if (metal_path == "") {
  metal_path <- possible_paths[file.exists(possible_paths)][1]
}
if (is.na(metal_path) || metal_path == "") stop("METAL binary not found. Check build logs in METAL/build.")

cat("METAL binary found at:", metal_path, "\n")
```


```{r}
metal_script <- c(
  "# METAL meta-analysis commands",
  "SCHEME STDERR",

  "# READ in FinnGen data",
  "MARKER SNP",
  "ALLELE A1 A2",
  "EFFECT B",
  "STDERR SE",
  "PVALUE P",
  "WEIGHT N",
  "PROCESS /content/chd_summaries.FinnGen.txt",

  "# READ in UK Biobank data",
  "MARKER rsid",
  "ALLELE effect_allele other_allele",
  "EFFECT beta",
  "STDERR standard_error",
  "PVALUE p_value",
  "WEIGHT n",
  "PROCESS /content/chd_summaries.UKBB.txt",

  "ANALYZE"
)
writeLines(metal_script, con = "metal_script.txt")
```

```{r}

log_out <- "metal_run.log"
log_err <- "metal_run.err"
cmd <- sprintf("%s metal_script.txt > %s 2> %s", shQuote(metal_path), log_out, log_err)
system(cmd, wait = TRUE)


cat("=== METAL STDOUT (tail) ===\n")
system(sprintf("tail -n 50 %s || true", log_out))
cat("\n=== METAL STDERR (tail) ===\n")
system(sprintf("tail -n 50 %s || true", log_err))


list.files(pattern = "META|meta_results|METAANALYSIS", recursive = FALSE)
```

```{r}
shel_call("mv METAANALYSIS1.TBL METAANALYSIS1_Septal.TBL")
```































